---
title: The Fall of an Empire
author: Emerson M. Del Ponte
date: '2019-05-25'
slug: the-fall-of-an-empire
categories: ["R", "Tutorial", "Programming", "Visualization"]
tags: [""]
header:
  caption: ''
  image: ''
output: html_document
---
I recently tweeted an animated GIF that depicted the yearly variation in efficacy of tebuconazole, a DMI fungicide, evaluated in field trials, for the control of soybean rust in a major soybean region of Brazil (Mato Grosso state). 

<blockquote class="twitter-tweet" align = center data-lang="pt"><p lang="en" dir="ltr">This is what happens when one fungicide (DMI in this case) is massively used over many years in a large soybean-growing region. Life finds a way. Showed this in my talk this AM &quot;To spray or not to spray&quot;. <a href="https://t.co/uFf6VbSieb">pic.twitter.com/uFf6VbSieb</a></p>&mdash; Emerson Del Ponte (@edelponte) <a href="https://twitter.com/edelponte/status/1131633376182579207?ref_src=twsrc%5Etfw">23 de maio de 2019</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

<br><br>
The Asian rust is the most damaging disease of soybean in Brazil since it was discovered in 2004, for which no other method than fungicides are available so far for an effective control. The tweet got a lot of attention, not only because of the important issue, but I think also because of the nice way to visualize the data. So I decided to share the R code use to produce the animation for other's presentations.

The data used here are from a more complete set obtained from a meta-regression study where the effect of year was tested as moderator variable ([Dalla Lana et al. 2018](https://apsjournals.apsnet.org/doi/abs/10.1094/PDIS-03-17-0408-RE)). The data of that study have been made available ([Del Ponte et al. 2017](https://osf.io/7d3ch/)), which makes it possible to reuse them!

We will start by importing, inspecting and preparing the data for visualization. I like the tidyverse tools and programming style, so this package is essentially the first to load. The data are available in one of my [OSF project](https://osf.io/7d3ch/) as an `csv` file. In the files panel, look for the `sev_data.csv` file within the data subfolder. Let's import the web-hosted file into R  using `read_csv` function. You need Internet connection, but you may want to save the data frame locally. 


```{r message=FALSE, warning=FALSE}
library(tidyverse)
fungicide <- read_csv("https://osf.io/rp7nb/download")
```


## Data inspection


Let's see how many treatments and the number of entries per treatment. These are the different levels of the `active_ingredient` column, including the non-treated check. I like to use the `tabyl` function of the *janitor* package as substitute for `table()`, so you can do the same using the latter base function if you prefer.



```{r}
library(janitor)
fungicide %>%
  tabyl(active_ingredient) %>%
  arrange(-n)
```

And now we can make a pivot table for active ingredient versus year to check those used for longer time and lower gaps. The values represent the number of trials where the fungicide was evaluated within a specific year.


```{r}
fungicide %>%
  tabyl(active_ingredient, year) %>%
  arrange(active_ingredient)
```

Besides the non-treated check with the largest sample size (250 independent experiments), the fungicides evaluated in most years were __tebu__ and __azox+cypro__, or eight years (2007 to 2014). We will need the check treatment as well. I like to subset the data using the `filter()` function so we keep only the relevant data to work further.


## Data preparation

```{r}
tebu_mix <- fungicide %>%
  filter(active_ingredient %in% c("check", "tebu", "azox_cypr")) %>%
  arrange(trial) # ordered by trial
```

The trials were conducted at several states in Brazil. Let's see how many trials per state and year.


```{r}
fungicide %>%
  tabyl(state, year)
```

I want to eliminate from our analysis states with too many gaps in the years and group states by climatic region or geographical proximity. I split all six remaining states into three groups as follows.

```{r}
tebu_mix1 <- tebu_mix %>%
  filter(!state %in% c("SC", "TO", "MA", "MG", "BA", "DF")) %>%
  mutate(region = case_when(
    state %in% c("MT", "GO") ~ "MT+GO",
    state %in% c("SP", "MS") ~ "SP+MS",
    state %in% c("PR", "RS") ~ "PR+RS"
  ))
```



Now we need to create our variable of interest, or the percent reduction in disease severity relative to the non-treated check, commonly referred to control efficacy. This can be done by first reshaping data to the wide format using the `spread()` function. Then, with the three response variables in separate columns, we can create the new variable of interest using the `mutate()` function. The efficacy is obtained from the ratio of fungicide-treated and non-treated subtracted from 1. Hence, the larger the efficacy (zero to 100%), the better is the fungicide. Note that `selection()` was used to preserve only the columns that are relevant. Note: negative values can be calculated in rare situations where disease severity in the fungicide-treated plot is higher than in the non-treated check of the same experiment.


```{r}
tebu_mix2 <- tebu_mix1 %>%
  select(trial, year, state, region, active_ingredient, severity) %>%
  spread(active_ingredient, severity) %>%
  mutate(
    tebu_eff = 100 - (tebu / check * 100),
    azox_cypr_eff = 100 - (azox_cypr / check * 100)
  ) %>%
  select(1:4, 8:9)
head(tebu_mix2)
```
 
## Data visualization

Now that our dataset is ready, let's make a static plot just for quick visualization. First we want to reshape the data back to the long format using `gather()` for producing side by side plots by another variable using `facet_wrap`.


```{r fig.height=5, fig.width=7}

# reorder levels of the region factor
tebu_mix2$region2 <- factor(tebu_mix2$region, levels = c("MT+GO", "SP+MS", "PR+RS"))

tebu_mix2 %>%
  select(year, region2, state, tebu_eff, azox_cypr_eff) %>%
  gather(key = treatment, value = efficacy, -c(1:3)) %>%
  ggplot(aes(year, efficacy, color = region2)) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  facet_grid(region2 ~ treatment) +
  theme_grey() +
  theme(legend.position = "top") +
  labs(x = "Harvest year", y = "Percent disease reduction", color = "Region")
```



## gganimate plots

Now the fun plots! we will use the _gganimate_ package to produce a GIF animation. There is no trouble here as we just need to add `year` variable in the `transition_time` function. Further details in official [package documentation](https://gganimate.com/articles/gganimate.html). I learnt that in my example year should be transformed to an integer to prevent it from display several decimals. 

Three other functions I found interesting to add to the animation (see after `transition_time`). More tricks on using gganimate [here](https://www.listendata.com/2019/05/create-animation-in-r-learn-with.html?m=1)



```{r echo=TRUE, message=TRUE, warning=TRUE}
library(gganimate)

tebu_decline <- tebu_mix2 %>%
  select(year, region2, tebu_eff, azox_cypr_eff) %>%
  gather(key = treatment, value = efficacy, -c(1:2)) %>%
  filter(efficacy > 0)

tebu_anima <- tebu_decline %>%
  ggplot(aes(year, efficacy, color = region2)) +
  geom_jitter(
    size = 3,
    alpha = 0.5
  ) + 
  scale_x_continuous(breaks = c(2005, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015)) +
  transition_time(as.integer(year)) +
  ease_aes("linear") +
  exit_fade() +
  shadow_trail(max_frames = 7) +
  labs(
    title = "Efficacy ot fungicides for soybean rust control",
    subtitle = "Each dot is one experiment. Year: {frame_time}",
    x = "Harvest year",
    y = "Disease reduction relative to non-treated (%)",
    caption = "Source: Dalla Lana (2018)",
    color = "States"
  ) +
  theme_minimal()+
  theme(legend.position = "right") +
  facet_grid(region2 ~ treatment)
animate(tebu_anima, width = 600, height = 500) 
```

This is it! Don't forget to save this current animation to GIF using:

```{r }
anim_save("tebuconazole-decline.gif", last_animation())
```

