---
title: Fitting Linear and Quadratic Models of Nematode Population Responses to time and Temperature
author: AH Sparks and JP Thompson
date: '2018-01-29'
slug: linear-models-nematode-time-temperature-response
categories:
  - modeling
tags:
  - teaching, nematodes, linear modeling
summary: 
output:
  blogdown::html_page:
    number_sections: false
  
---


# Introduction

_Pratylenchus thornei_, the root-lesion nematode is widely distributed in wheat
(_Triticum aestivum_) growing areas of many countries and is of particular
concern in sub-tropical environments (Thompson 2015). These nematodes penetrate
roots to  feed and reproduce in the root cortex leading to loss of root
function, which affects nutrient and water uptake of nutrients and water causing
nutrient deficiency and water stress (Thompson 2015).

In the original paper the population of _P. thornei_ in wheat in Queensland,
Australia is modelled using a linear and quadratic equations. The study aimed to
investigate the effects of soil profile temperatures after different sowing
dates on reproduction of the nematodes in susceptible and moderately resistant
wheat cultivars in the subtropical grain region of eastern Australia. This
document recreates the models for population densities of _P. thornei_ as
described in _Modelling population densities of root-lesion nematode_ (P.
thornei) _from soil profile temperatures to choose an optimum sowing date for
wheat_ (Thompson 2015).

There are two types of models described in the paper, the first model is a
linear model used to describe the unplanted control and two quadratic models fit
Gatcher (Susceptible) and GS50a (Moderately Resistant) wheat cultivars.

## Session Setup

Using the _tidyverse_ package simplifies the libraries. It loads, _readr_, used
to import the data, _tidyr_ used to format the data and _ggplot2_ used for
visualising the data.

```{r libraries_and_seed, message=FALSE}
library(tidyverse)
```

We will use the `set.seed` function for reproducibility.

```
set.seed(52)
```

## Data Import and Inspection

Import the data using `read.csv` from _readr_, which is a part of the
_tidyverse_.

```{r data_import, echo=TRUE, message=FALSE}
nema <- read_csv("data/Degree Days Relationships.csv")

nema
```

You can see that each of the varieties have their own column in the original
data format (wide). Using `gather()` from the _tidyr_ package (part of the
_tidyverse_), convert from to long format where the varieties are all listed in
a single column, "Variety". The `data` paramter tells R which data frame to
gather. The `key` parameter is the name of the new column to be created called
"Variety", `value` specifies the column that will contain the values that go with
the varieties, "Population". The last portion tells `gather()` which columns are
to be gathered. Using the ":" means take the columns from "Unplanted" to
"Suneca" and gather them without needing to type all the column names.

```{r gather_data, echo=TRUE}
nema_long <-
  gather(data = nema,
         key = Variety,
         value = Population,
         Unplanted:Suneca)

nema_long
```

Now that the data are in the format that _ggplot2_ prefers, take a look at the
data first to see what it looks like. Fit a smoothed line for each Variety to
the raw data. The individual temperatures are shown here by shape, the variety
by colour.

```{r raw_data_scatterplots, echo=TRUE}
ggplot(
  nema_long,
  aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )
) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth() +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5))
```

# Model Fitting

## Unplanted Model

The paper uses a linear model for the unplanted control. Here write a function
to use in modelling the unplanted population data.

### Linear Model

```{r linear_model_HT, echo=TRUE}
linear_model <- function(df) {
  lm(Population ~ Degree_days,
     data = df)
}
```

Now check the model fit, using `filter()` from _dplyr_ to select only Unplanted
data from the data set for the model and fit the linear model to the data.

Lastly we can show the unplanted data alone as a scatterplot with the model
line fitted using `geom_smooth()` from _ggplot2_.

```{r check_model}
unplanted_model <- nema_long %>%
  filter(Variety == "Unplanted") %>%
  linear_model()

par(mfrow = c(2, 2))
plot(unplanted_model)

summary(unplanted_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Unplanted") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("Unplanted Linear Model")
```

## Quadratic Models

In the original paper, Gatcher and GS50a best fit quadratic models, which are
fit here.

```{r HT_quadratic_models, echo=TRUE}
quadratic_model <- function(df) {
  lm(Population ~ Degree_days + I(Degree_days^2),
      data = df)
}
```

### Susceptible Varieties

Gatcher, Potam and Suneca all have very similar curves, here Gatcher is used to
fit a quadratic model as in the original paper following the same methods as
above for the linear model.

```{r susceptible_model, echo=TRUE}
s_model <- nema_long %>%
  filter(Variety == "Gatcher") %>% 
  quadratic_model()

par(mfrow = c(2, 2))
plot(s_model)

summary(s_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "Gatcher") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("Gatcher Quadratic Model")
```

### Moderately Resistant Cultiver

GS50a, moderately resistant to _P. thornei_ also fit a quadratic model but the
coefficients were slightly different due to different responses to the variety
and temperature.

```{r moderately_resistant_model, echo=TRUE}
mr_model <- nema_long %>%
  filter(Variety == "GS50a") %>%
  quadratic_model()

par(mfrow = c(2, 2))
plot(mr_model)

summary(mr_model)

nema_long %>%
  group_by(Variety) %>%
  filter(Variety == "GS50a") %>%
  ggplot(aes(
    x = Degree_days,
    y = Population,
    colour = Variety,
    group = Variety
  )) +
  geom_point(aes(shape = as.factor(Temperature))) +
  geom_smooth(method = "lm",
              formula = y ~ x + I(x^2),
              size = 1) +
  ylab(expression(paste("ln(",
                        italic("P. thornei"),
                        "/kg soil) + 1"),
                  sep = "")) +
  xlab("Thermal Time (˚C Days Above 10˚C)") +
  theme_minimal() +
  scale_shape_discrete("Temperature") +
  scale_colour_discrete("Variety") +
  theme(axis.text.x  = element_text(angle = 90,
                                    vjust = 0.5)) +
  ggtitle("GS50a Quadratic Model")
```

As in the original paper, the model equations can be derived from these models
as well.

# Reproducibility
```{r reproducibility, echo=FALSE}
devtools::session_info()
```
