---
title: "Calculating AUDPS using Julia"
description: |
  Calculating AUDPS in R and Julia.
author:
  - name: Adam Sparks
    url: https://adamhsparks.netlify.app
date: 2022-02-13
bibliography: bibliography.bib
output:
  distill::distill_article:
    self_contained: false
preview: Julia_Programming_Language_Logo.svg
csl: phytopathology.csl
categories: 
  - Julia4PlantPath
  - R4PlantPath
  - Reproducible Research
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

![](Julia_Programming_Language_Logo.svg)

Previously we introduced Julia, a programming language that is similar to R or Python and demonstrated how AUDPC can be calculated using the trapezoidal method in R as shown in the "Disease Progress Over Time" module of the "Epidemiology and Ecology in R", Sparks et al. [-@Sparks2008].
Then we looked at how the function could be optimised in R before writing a Julia function to calculate the same value.

Now we will take a look at a similar calculation, Area Under the Disease Stairs (AUDPS) [@Simko2012].
AUDPS can give better estimates of the disease progress by giving a weight closer to the optimal first and last observations.

This function is not the fully optimised version like what we showed for AUDPC, using `sum()` would help make this faster but possibly at the expense of readability so we'll stick with using the regular `+` and `-` here for readability.

```{r, echo=TRUE}
r_audps <-
  function(evaluation, dates) {
    n <- length(dates)
    n_1 <- length(dates) - 1
    
    x <- 0
    for (i in seq_len(n_1)) {
      x <- x + evaluation[[i]] * (dates[[i + 1]] - dates[[i]])
    }
    audps <- x + evaluation[[n]] * (dates[[n]] - dates[[n_1]])
    return(audps)
  }
```

Both of the R packages that were previously discussed when showing how to calculate AUDPC, [_agricolae_](https://cran.r-project.org/package=agricolae) [@deMendiburu2021] and [_epifitter_](https://cran.r-project.org/package=epifitter) [@Alves2021], provide easy to use functions to calculate AUDPS, `audps()` and `AUDPS()`, respectively.
The following code uses an example from _agricolae's_ help showing how to calculate AUDPS in R using our own function, `r_audps()`.

```r
> dates <- c(14, 21, 28) # days
> evaluation <- c(40, 80, 90)
> r_audps(evaluation, dates)
```

```{r, echo=FALSE}
dates <- c(14, 21, 28) # days
dates
# example 1: evaluation - vector
evaluation <- c(40, 80, 90)
evaluation
r_audps(evaluation, dates)
```

### Using Julia

Since we've already introduced Julia, here we'll just build an AUDPS function in Julia to illustrate how it can be done.

```{julia audps, echo=TRUE}
function j_audps(evaluation, dates)
    
    # find how many observations there are and calculate that minus 1 as well
    n = length(dates)
    n_1 = length(dates) - 1

    # initialise our objects outside the loop
    i = 0
    out = 0
    # the for loop looks roughly the same but here we just use 1:n
    for j in 1:n_1
        i = i + evaluation[j] * (dates[j + 1] - dates[j])
        
        out = i + evaluation[n] * (dates[n] - dates[n_1])
    end
  
    # return the object, `out` from the for loop
    return out

# end the function (no curly brackets!)
end
```

```julia
julia> dates = [14, 21, 28] # days
julia> evaulation = [40, 80, 90]
julia> j_audps(evaulation, dates)
```

```{julia run-audps, echo=FALSE}
dates = [14, 21, 28] # days
evaulation = [40, 80, 90]
j_audps(evaulation, dates)
```

The AUDPS values match!

## Conclusion

This is just a quick follow-up example of how you can use Julia in plant pathology to show new users how it compares with R with a another commonly used function.
If you're curious to know more, the [Julia docs](https://docs.julialang.org) are a great place to start.
In particular, the [noteworthy differences](https://docs.julialang.org/en/v1/manual/noteworthy-differences/#Noteworthy-differences-from-R) is a useful bit to refer to if you're familiar with R.

For a more detailed comparison of complete Julia and R packages that offer an existing plant disease model, EPIRICE [@Savary2012], see [Epicrop.jl](https://adamhsparks.github.io/Epicrop.jl/stable/) [@Epicrop.jl-2022], a port of [epicrop](https://adamhsparks.github.io/epicrop/) [@Sparks2021] to Julia, which has demonstrated faster speeds in benchmarking tests for the same rice disease predictions.

## Colophon

This post was constructed using R Version ```r paste(R.Version()[c("major", "minor")], collapse = ".")``` [@RCore2021] and Julia Version ```r JuliaCall::julia_eval("VERSION")``` [@Bezanson2017] using _JuliaCall_ Pull Request [#174](https://github.com/Non-Contradiction/JuliaCall/pull/174).
