---
title: "Using technilogy to advance the the sience: Digitizing  research results to produce crop disease risk prediction tools"
author:
- name: Mladen Cucak
  affiliation: Maynooth University Department of Geography, Maynooth University, Co. Kildare, Ireland; Teagasc Crops Research Centre, Oak Park, Carlow, Ireland
- name: Neil McRoberts
  affiliation: Quantitative Biology and Epidemiology (QBE) lab, Plant Pathology Department at UC Davis
output:
  blogdown::html_page:
    highlight: tango
    df_print: paged
bibliography: bibliography.bib
csl: phytopathology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)

# check for klippy and remotes, install as necessary
if (!require("klippy")) {
  if (!require("remotes")) {
  install.packages("remotes", repos = "http://cran.rstudio.com/")
  }
  remotes::install_github("RLesur/klippy")
}

klippy::klippy(position = c('bottom', 'left'))
```

# Using 

## Introduction

Text[@rotem_relativity_1971].

## Packages
```{r libraries, message=FALSE}
list.of.packages <-
  c(
    "readxl",
    "here", #package that helps paths in different platforms
    "stringr", #string manipulations
    "reshape2", #Reshaping the data to long format, needed for plotting and model fitting
    "egg", #Plotting aids: interesting themes and putting together a number of plots
    "RColorBrewer", #Color palette
    "mgsub"

  )

new.packages <-
  list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]

#Download packages that are not already present in the library
if (length(new.packages))
  install.packages(new.packages)

packages_load <-
  lapply(list.of.packages, require, character.only = TRUE)

#Print warning if there is a problem with installing/loading some of packages
if (any(as.numeric(packages_load) == 0)) {
  warning(paste("Package/s", paste(list.of.packages[packages_load != TRUE]), "not loaded!"))
} else {
  print("All packages were successfully loaded.")
}
```

# Data

!!!!Check stadnard errors for various infection levels 

```{r}
dis_df <- read_excel(here::here("cucak", "1", "dat", "P_inf_infection_Rotem.xlsx"))
dis_df <-  
  reshape2::melt(dis_df, id.var = 1:2, variable.name = "temp", value.name = "dis_inc",factorsAsStrings = F)
dis_df$temp <- as.numeric(as.character.factor(dis_df$temp))
dis_df$wet_dur_lab <- factor(dis_df$wet_dur)
levels(dis_df$wet_dur_lab) <- paste0(levels(dis_df$wet_dur_lab), "h of s. wetness") 

dis_df$inoculum_dose <- factor(dis_df$inoculum_dose, ordered = T)
head(dis_df)
```


```{r}
ggplot(dis_df,aes(x= temp, y = dis_inc, group = inoculum_dose, colour = inoculum_dose))+
geom_point()+
geom_line()+
facet_wrap(wet_dur_lab~., ncol = 4)+
scale_color_brewer(palette="GrandBudapest")+
labs(
title = "",
x = "Temperature (°C)",
y = "Disease Incidence (grades)",
color = expression(paste("Inoculum dose ( Sporangia", "/",cm^2,")", sep="")),
caption = expression(paste(
"\nFigure 1. The effect of temperature (T, in °C), inoculum dose (Sporangia, in cm) and duration of wet period (W, in hr) on infection of potato leaves by ", "Phytopthora infestans: infection levels obtained for all treatment combinations (averaes of three replicate plants)."
)))+
theme_article()+
theme(legend.position = "top")
dis_df$wet_dur_lab <- NULL 
```


```{r}

dis_df$temp_lab <- factor(dis_df$temp)
levels(dis_df$temp_lab) <- paste0(levels(dis_df$temp_lab), "(°C)") 


ggplot(dis_df,aes(x= wet_dur, y = dis_inc, group = inoculum_dose, colour = inoculum_dose))+ 
  geom_point()+
  geom_line()+
  facet_wrap(temp_lab~., ncol = 6)+
  scale_color_brewer(palette="GrandBudapest")+
  scale_x_discrete(limits= unique(dis_df$wet_dur), labels = as.character(unique(dis_df$wet_dur)))+
  xlab("Wetness Duration (hours)")+ 
  ylab("Disease Incidence")+
  theme_classic()
```


We are trying to model a complex response to 3 interacting explanatory variables.
The Weibull function might be an appropriate model for the response to leaf wetness duration for two reasons.  First, it has a suitable sigmoid cumulative distribution and second it has a location (or shift) parameter that can be interpreted in biological terms as a minimum requirement for suitable conditions before a process will occur.  The Weibull function has many different parameterizations but one that is relevant for this problem is:$$
f(d) =\frac{\beta}{\eta} \left(\frac{t-\gamma}{\eta}\right)^{\beta-1} e^{\left(\frac{t-\gamma}{\eta}\right)^{\beta}} $$ 

Where $f(d)$ is the wetness-duration-dependent infection function, $\beta$ is the Weibull rate parameter, $\eta$ is the scale parameter (setting the characteristic duration over which events are completed), and $\gamma$ is the location parameter.

We choose a value (for function) for $\gamma$ that replicates what is know about the biology of the process.  This leaves the issue of the remaining parameters $\beta$ and $\eta$.   In essence, we construct functions for the response of these parameters to temperature and inoculum density, either from first principles, from data fitting, or a hybrid process.  Then we plug the derived functions into the formula for the Weibull function, which becomes a translation device for turning input data for temperature, inoculum level and wetness duration into values for infection intensity.

Writing generic functional forms for the functions for $\beta$ and $\eta$ as $\beta\equiv h(T,I)$ and $\eta\equiv l(T,I)$, where $T$ and $I$ are temperature and inoculum level, respectively, we obtain:$$
f(d) =\frac{h(T,I)}{l(T,I)} \left(\frac{t-\gamma}{l(T,I)}\right)^{h(T,I)-1} e^{\left(\frac{t-\gamma}{l(T,I)}\right)^{h(T,I)}} $$ 



# fitting the model

```{r}
dis_df$inoculum_dose <- as.numeric(as.character(dis_df$inoculum_dose))

poly.fit <- lm(dis_inc ~ poly(temp, 
                              wet_dur, 
                              inoculum_dose, 
                              degree = 3), 
               data = dis_df )

summary(poly.fit)
anova(poly.fit)

library(ggfortify)
autoplot(poly.fit, label.size = 2,
         data = dis_df, 
         colour = factor(dis_df$temp, ordered = T))+
  theme_article()

autoplot(poly.fit, which = 1:6, colour = 'dodgerblue3',
         smooth.colour = 'black', smooth.linetype = 'dashed',
         ad.colour = 'blue',
         label.size = 3, label.n = 5, label.colour = 'blue',
         ncol = 3)
#Problem scale location

```

# Reproducibility

```{r reproducibility, echo=FALSE}
devtools::session_info()
```

# References
